Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2008-12-01 docs/pubs/ISEI6/main.tex
abstract and introduction complete draft

\newcommand{\owlifier}{\textsf{Owlifier}}
\newcommand{\mypara}[1]{\vspace{12pt}\noindent\textbf{#1}}

@@

\begin{abstract}
  Data discovery and integration is important in many ecological
  studies, especially those that concern broad-scale ecological
  questions. While important, the discovery and integration of
  ecological data is often a complex and time-consuming task for
  researchers. This situation is often due to the use of informal,
  ambiguous, and sometimes inconsistent terms for describing
  ecological data.  Ontologies can help address this problem by
  providing a mechanism to formally define ecological concepts that
  can be used to more consistently annotate, relate, and search for
  data sets.

  Unlike in molecular biology or biomedicine, however, few ontology
  development efforts exist within ecology. These efforts often
  require considerable expertise in ontology languages and development
  tools that limits the creation of ontologies by ecologists.  This
  paper addresses this problem by allowing ecologists to use common
  spreadsheet tools to describe different aspects of an ontology. We
  introduce novel conventions for creating, relating, and constraining
  concepts through spreadsheets, and provide software tools for
  converting ontologies created in this way into equivalent OWL-DL
  representations. We also consider inverse translations, i.e., that
  convert ontologies represented using OWL-DL into our spreadsheet
  format. Our approach allows large lists of terms to be easily
  related and organized into concept hierarchies, and generally
  provides a more intuitive and natural interface for ontology
  creation by ecologists.
\end{abstract}

@@

Within the fields of molecular biology and biomedicine considerable
effort has gone into developing ontologies for improving data
discovery and integration
\cite{bard04:_ontol_in_biolog,ashburner00:_gene_ontol}. While similar
benefits can be obtained for ecological data, far fewer efforts exist
to develop richer and more consistent terminologies of ecology
concepts
\cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
The use of formal ontologies can significantly enhance metadata
descriptions of ecological data. Annotating data with ontology terms
can both help users interpret data as well as enable advanced
capabilities for data discovery and integration, e.g., by exploiting
subsumption and part-of hierarchies as well as more formal constraints
such as cardinality restrictions on properties and term
equivalence. Many efforts are underway to engage scientists in the
development of ontologies to represent common ecological and
biodiversity concepts. Most of these projects leverage the W3C Web
Ontology Language (OWL) \cite{smith04:_owl_web_ontol_languag_guide} as
a standard XML syntax for representing and sharing ontologies. A key
advantage of OWL is that it is supported by a wide range of generic
tools, including editors
\cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\cite{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\cite{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\cite{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
however, are primarily targeted at experts in knowledge engineering
and software development. This is especially true with ontology
editors (such as Protege, SWOOP, etc.), which allow for very detailed
ontology specifications, but at the same time require a considerable
amount of understanding of the underlying ontology formalisms and
syntax. Thus, we see the lack of suitable ontology editing tools for
scientists as one of the major barriers for more wide-scale adoption
of ontologies in ecology.

Here, we present a novel approach for ontology creation that aims at
being more intuitive for ecologists and that can be used to rapidly
construct large ontologies for describing scientific data. Our
approach is to allow scientists to use common spreadsheet-based tools
to describe, in an intuitive way, different aspects of an ontology,
and then to take these descriptions and convert them into full-fledged
OWL ontologies using a software application called \owlifier.

An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
a predefined template structure for users to fill in. Each non-empty
row in an owlifier table constitutes a block, where blocks can be used
to define ontology classes (including subclasses), class synonyms,
data- and object-type properties (and constraints), etc.  We also
provide blocks for plain-text descriptions of concepts and properties,
and for referencing one or more existing ontologies (e.g., to extend
an existing ontology or to define ontology articulations). Blocks can
be sparse (inheriting from previous blocks), which simplifies the
creation of large ontologies.

Although not as expressive as full-fledged OWL ontologies, our
approach can produce ontology structures essential for enhanced data
discovery and integration, while at the same time providing a more
accessible user interface for ecologists. Further, compared to many
existing editors, our approach can be used to rapidly construct
concept hierarchies, e.g., from long lists of keywords, within
familiar spreadsheet software. We have field-tested the owlifier
approach with ecologists and evolutionary biologists working with
trait data, and found this application enabled them to quickly and
easily comprehend and construct ontologies.

In this paper, we describe the syntax and semantics of \owlifier. We
consider a set of blocks that generally follow the ...

Notes:
\begin{itemize}
\item we generally follow the guidelines in ontology creation of
  Rector et al. in \cite{rector04:_owl_pizzas}.
\item we also try to simplify certain aspects, e.g., disjoint classes
  and open-world reasoning, which can be confusing for non-experts
\item we try to use more meaningful terms, when applicable (e.g.,
  instead of datatype property, attribute, etc.)
\item our goal is to not support all constructs in OWL-DL, but to
  provide the most commonly used ones for defining typical ontologies,
  but still allow experts to refine the ontology in ontology editing
  tools if necessary
\end{itemize}

\section{\owlifier\ Syntax and Semantics}

@@

\mypara{Concept Block.} Concept blocks introduce new OWL classes and
specify subclass relationships. Imported classes may also be used
within a concept block by prefixing the class name with a namespace
label, e.g., when an ontology is being extended.  We use the term
`concept' to avoid confusion with the use of the term 'class' in
biological taxonomies and because `concept' is often more intuitive
for users. Concept blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{concept} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each class $C_i$ is asserted in the current ontology to subsume
a class $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in a concept block
induces the DL axiom \[C_i \sqsubseteq C_{i+1}.\] If both $C_i$ and
$C_{i+1}$ are imported classes, we say that the block defines an
``articulation'' (i.e., mapping) between the two classes. The
following concept block defines a simple subclass hierarchy
\begin{tabbing}
  ~~\texttt{concept PhysicalFeature AquaticPhysicalFeature River}
\end{tabbing}
which states that river is a subclass of aquatic physical feature and
that aquatic physical feature is a subclass of physical feature.

>>>>

\newcommand{\Owlifier}{\textsf{Owlifier}}
\newcommand{\owlifier}{\textsf{owlifier}}
\newcommand{\mypara}[1]{\vspace{12pt}\noindent\textbf{#1}}

\newcommand{\secref}[1]{Sect.~\ref{#1}}
\newcommand{\figref}[1]{Fig.~\ref{#1}}

@@

\begin{abstract}
  Discovery and integration of data is important in many ecological
  studies, especially those that concern broad-scale ecological
  questions. Discovery and integration is often a difficult and
  time-consuming task for researchers, in part because informal,
  ambiguous, and sometimes inconsistent terms are used to describe
  data semantics.  Ontologies can help address this problem by
  providing a means to define ecological concepts to more consistently
  annotate, relate, and search for data sets.
  %
  However, unlike in molecular biology or biomedicine, few ontology
  development efforts exist within ecology. Ontology development often
  requires considerable expertise in ontology languages and
  development tools, which is often a barrier for ontology creation in
  ecology. In this paper we address this problem by providing an
  approach for ontology creation that allows ecologists to use common
  spreadsheet tools to describe different aspects of an ontology.  We
  present conventions for creating, relating, and constraining
  concepts through spreadsheets, and provide software tools for
  converting these ontologies into equivalent OWL-DL
  representations. We also consider inverse translations, i.e., to
  convert ontologies represented using OWL-DL into our spreadsheet
  format. Our approach allows large lists of terms to be easily
  related and organized into concept hierarchies, and generally
   provides a more intuitive and natural interface for ontology
  development by ecologists.
\end{abstract}

@@

Within the fields of molecular biology and biomedicine considerable
effort has gone into developing ontologies for improving data
discovery and integration
\cite{bard04:_ontol_in_biolog,ashburner00:_gene_ontol}. While similar
benefits can be obtained for ecological data, far fewer efforts exist
to develop richer and more consistent terminologies of ecology
concepts
\cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
The use of formal ontologies can significantly enhance metadata
descriptions of ecological data. Annotating data with ontology terms
can both help users interpret data as well as enable advanced
capabilities for data discovery and integration, e.g., by exploiting
subsumption and part-of hierarchies as well as more formal constraints
such as cardinality restrictions on properties and term equivalence.

Efforts to engage scientists in the development of ontologies
typically leverage the W3C Web Ontology Language (OWL)
\cite{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
for representing and sharing ontologies. A key advantage of OWL is
that it is supported by a wide range of generic tools, including
editors
\cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\cite{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\cite{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\cite{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
however, are primarily targeted at experts in knowledge engineering
and software development familiar with the underlying description
logic semantics of OWL-DL \cite{grau08:_owl}. This is especially true
with ontology editors (such as Protege, SWOOP, etc.), which allow for
very detailed ontology specifications, but at the same time require a
considerable amount of understanding of the underlying ontology
formalisms and syntax. Thus, we see the lack of suitable ontology
editing tools for scientists as one of the major barriers for more
wide-scale adoption of ontologies in ecology.

This paper presents a novel approach for ontology creation that aims
at being more intuitive for ecologists and that can be used to rapidly
construct large ontologies for describing scientific data. Our
approach is to allow scientists to use common spreadsheet-based tools
to describe, in an intuitive way, different aspects of an ontology,
and then to take these descriptions and convert them into full-fledged
OWL ontologies using a software application called \owlifier. An
\owlifier\ spreadsheet consists of a set of \emph{blocks} that have a
predefined template structure for users to fill in. Each non-empty row
in an \owlifier\ table constitutes a block. Each block defines
different aspects of an ontology including ontology classes,
subclasses, synonyms, and properties.  We also provide blocks for
plain-text descriptions of classes and properties, and for referencing
one or more existing ontologies (e.g., to extend an existing ontology
or to define ontology articulations). Blocks can be sparse (inheriting
from previous blocks), which can further simplify the creation of
large ontologies.

While not as expressive as OWL-DL, our approach can produce ontology
structures essential for enhanced data discovery and integration,
while at the same time provide a more accessible user interface for
ecologists. Further, our approach can be used to rapidly construct
class hierarchies from long lists of keywords using familiar
spreadsheet software. For instance, an ecologist can easily list (or
import) a set of terms, and then incrementally organize these into a
class hierarchy, define properties and constraints, etc. We have used
the \owlifier\ approach with ecologists and evolutionary biologists
working with trait data, and found that it enabled them to quickly and
easily comprehend and construct ontologies.

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. We define blocks that support a large subset of OWL-DL and
that also generally follow the ontology creation guidelines defined in
\cite{rector04:_owl_pizzas}. We also simplify certain aspects of
ontology creation using OWL-DL, e.g., by assuming classes are disjoint
by default (unless specified otherwise) and by applying implicit
property restriction closures \cite{rector04:_owl_pizzas}. In
\secref{sec:properties} we describe additional characteristics of
\owlifier\ and discuss issues with respect to classification and
reasoning. In \secref{sec:implementation} we briefly describe the
\owlifier\ implementation, and conlcude in \secref{sec:conclusion}
with related and future work. In general, the goal of \owlifier\ is
not to support all constructs in OWL-DL, but instead to provide a
higher-level ontology syntax (via spreadsheet blocks) that is easy for
ecologists to use and understand while also providing the necessary
constructs for developing typical ecological ontologies. By compiling
\owlifier\ to OWL-DL, we also allow for experts to refine and extend
the ontology using more advanced ontology editing tools if necessary.

\section{\Owlifier\ Syntax and Semantics}
\label{sec:owlifier}

@@

\mypara{Entity Block.} Entity blocks introduce new OWL classes and
specify subclass relationships. Imported classes may also be used
within a concept block by prefixing the class name with a namespace
label, e.g., when an ontology is being extended.  We use the term
`entity to avoid confusion with the use of the term 'class' in
biological taxonomies and because `entity' is used in \cite{oboe} to
define ecological concepts. Entity blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \texttt{entity} & $C_1$ & $C_2$ & \dots & $C_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each class $C_i$ is asserted in the current ontology to subsume
a class $C_{i+1}$, for $1 \le i < n$. Each $C_i$ in a concept block
induces the DL axiom \[C_i \sqsubseteq C_{i+1}.\] If both $C_i$ and
$C_{i+1}$ are imported classes, we say that the block defines an
``articulation'' (i.e., mapping) between the two classes. The
following concept block defines a simple subclass hierarchy
\begin{tabbing}
  ~~\texttt{entity PhysicalFeature AquaticPhysicalFeature River}
\end{tabbing}
which states that river is a subclass of aquatic physical feature and
that aquatic physical feature is a subclass of physical feature.
